---
title: "Exploring Wage Disparities in California, within the Education Sector, and within Colleges and Universities"
author: "Audrey Omidsalar"
date: "December 10th, 2021"
output:
  html_document:
      toc: TRUE
      toc_float: TRUE
always_allow_html: true
---

A pdf version of the report can be found [here](https://github.com/aomidsalar/PM566_Final_Project/blob/main/index.pdf)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('data.table')
library('ggplot2')
library('dplyr')
library('tibble')
library('tidyverse')
library('plotly')
library('DT')
library('knitr')
library('priceR')
```


```{r download, cache = TRUE, echo=FALSE}
if (!file.exists("Quarterly_Census_of_Employment_and_Wages__QCEW_.csv"))
  download.file(
    url = "https://data.edd.ca.gov/api/views/fisq-v939/rows.csv?accessType=DOWNLOAD",
    destfile = "Quarterly_Census_of_Employment_and_Wages__QCEW_.csv",
    method   = "libcurl",
    timeout  = 60
    )
input <- data.table::fread("Quarterly_Census_of_Employment_and_Wages__QCEW_.csv")
```

# Introduction
  An issue that I have seen being constantly spoken about is the economy and the rise of inflation. Being a student in the science field, I have very little background on economics; frankly, the last time I took a course in economics was in high school. I was interested in examining wage disparities, more specifically throughout California, within the education sector and within colleges and universities, as this is the area I'm currently working in.

#### The question that I will be exploring is: How have wages throughout California, in the education sector, and in Colleges and Universities changed from 2004 until 2020?

  The dataset I used is California's Quarterly Census of Employment and Wages from data.gov (https://catalog.data.gov/dataset/quarterly-census-of-employment-and-wages-qcew). 

The specifications I was interested in looking in for this project were the average weekly wages throughout various counties in California, within the Education as a whole, and within the *Colleges and Universities* industry.

# Methods

The dataset was cleaned according to the exploratory data analysis checklist and was subset to make a smaller datasets encompassing the *education*, *colleges and universities*, *privately owned colleges and universities*, and *state government owned colleges and universities* categories. The average weekly wages for colleges and universities were adjusted for inflation. More detailed information about these methods can be found in the [pdf report](https://github.com/aomidsalar/PM566_Final_Project/blob/main/index.pdf).

The graphs in this report were made using the ggplot2 package, and interactive plots were made with plotly.

```{r filtering, echo=FALSE}
#step 1
input <- filter(input, Quarter == "Annual")
#step 2
input <- filter(input, `Area Type` != "United States")
input = subset(input, select = -c(`1st Month Emp`,`2nd Month Emp`, `3rd Month Emp`) )
#step 3
education <- input %>% filter(str_detect(`NAICS Code`, '^61'))
```
# Results

### 1. Which industries have the highest and lowest wages?

  To do this, I first searched for the median average weekly wage for each unique industry overall per year. I chose to look at the median rather than the mean in order to control for any outliers. There are likely industries that have seasonal changes, and I felt the median would be a fair estimation to compare. The top and bottom five industries are listed below.

```{r industries, echo=FALSE}
industries <- input[, .(
  median_wage = median(`Average Weekly Wages`, na.rm = TRUE)), 
  by = .(`Industry Name`, Year)
]
```


```{r industries top, echo=FALSE}
industries[order(-industries$median_wage), ] %>% head(5) %>% select(`Industry Name`, `Year`, median_wage) %>% knitr::kable(caption = "5 Industries with the Highest Median Weekly Wages per Year")
```


```{r industries bottom, echo=FALSE}
industries[order(industries$median_wage), ] %>% head(5) %>% select(`Industry Name`, `Year`, median_wage) %>% knitr::kable(caption = "5 Industries with the Lowest Median Weekly Wages per Year")
```

### 2. Which counties have the highest and lowest wages?

  To do this, I found the median of the average weekly wage for each unique county per year. Again, I chose to look at the median rather than the mean in order to control for any outliers. The top and bottom five industries are listed below.

```{r counties, echo=FALSE}
counties <- input[, .(
  median_wage = median(`Average Weekly Wages`, na.rm = TRUE)), 
  by = .(`Area Name`, Year)
]
```

  
```{r counties top, echo=FALSE}
counties[order(-counties$median_wage), ] %>% head(5) %>% select(`Area Name`, `Year`, median_wage) %>% knitr::kable(caption = "5 Counties with the Highest Median Weekly Wages per Year")
```

  
```{r counties bottom, echo=FALSE}
counties[order(counties$median_wage), ] %>% head(5) %>% select(`Area Name`, `Year`, median_wage) %>% knitr::kable(caption = "5 Counties with the Lowest Median Weekly Wages per Year")
```

### 3. Which education sectors have the highest and lowest weekly wages?

  I filtered the dataset to the top average weekly wages per industry. The top and bottom five are listed below.
  
```{r edsector, echo=FALSE}
edsector <- education[, .(
  median_wage = median(`Average Weekly Wages`, na.rm = TRUE)), 
  by = .(`Area Name`, Year, `Industry Name`)
]
```

  
```{r edsector top, echo=FALSE}
edsector[order(-edsector$median_wage), ] %>% head(5) %>% select(`Area Name`, `Industry Name`, `Year`, median_wage) %>% knitr::kable(caption = "Top 5 Median Weekly Wages per Year")
```

  
```{r edsector bottom, echo=FALSE}
edsector[order(edsector$median_wage), ] %>% head(5) %>% select(`Area Name`, `Industry Name`, `Year`, median_wage) %>% knitr::kable(caption = "Bottom 5 Median Weekly Wages per Year")
```

  
```{r topedsector, echo=FALSE}
topedsector <- education %>% group_by(`Industry Name`) %>% top_n(1, `Average Weekly Wages`) %>% select(`Area Name`, `Industry Name`, `Year`, `Ownership`, `Establishments`,`Average Monthly Employment`,`Total Wages (All Workers)`,`Average Weekly Wages`)
#topedsector[order(-topedsector$`Average Weekly Wages`), ] %>% head(5) %>% select(`Area Name`, `Industry Name`, `Year`, `Average Weekly Wages`) %>% knitr::kable(caption = "5 Education Sectors with the Highest Average Weekly Wages")
datatable(topedsector, caption="Education Sectors with the Highest Average Weekly Wages")
```

### 4. Within the *Colleges and Universities* sector, how have wages changed from 2004 to 2020?

**In order for the comparisons on wages to be a little more fair, I adjusted these wages for inflation, using the `priceR` package. I corrected the values in the *Average Weekly Wages* column for inflation in the US to the year 2020 (this is the most recent year in this colleges and universities dataset). All subsequent data and graphs will be based on these inflation-corrected wages.**

  This first scatterplot shows the annual average weekly wages in local government, private, and state government colleges and universities throughout California. The outliers that pop out to me are private colleges/universities in San Mateo County (Stanford is located here), Santa Clara County (Santa Clara University is here), and Los Angeles County (USC, among others). These are all in areas with major cities and a high cost-of-living.
  
  Additionally, there seems to generally be a smaller range in average wages in the colleges and universities that are owned by the state (the points are more clustered together vertically), with San Joaquin County being an exception with comparatively lower wages. This is likely due to standards and policies that are maintained across these institutions.
  
```{r inflation, echo=FALSE, message=FALSE, results=FALSE}
colors <- c("Alameda County" = "#8B0000", "Contra Costa County" = "#CD5C5C", "Fresno County" = "#FFC0CB", "Kern County" = "#FFA500", "Los Angeles County" = "#FF4500", "Monterey County" = "#FFD700", "Orange County" = "#BDB76B", "Riverside County" = "#3CB371", "Sacramento County" = "#006400", "San Bernardino County" = "#7CFC00", "San Diego County" = "#20B2AA", "San Francisco County" = "#AFEEEE", "San Luis Obispo County" = "#5F9EA0", "Santa Barbara County" = "#1E90FF", "Santa Clara County" = "#0000CD", "Solano County" = "#9370DB", "Sonoma County" = "#800080", "San Mateo County" = "#FF00FF", "Ventura County" = "#808080", "Stanislaus County" = "#D3D3D3", "Tulare County" = "#FF0000", "San Joaquin County" = "#7DF9FF")
colluniv <- filter(education, `Industry Name` == "Colleges and Universities")
#p1 <- colluniv[`Area Name` != "California"] %>% ggplot() +
#  geom_point(mapping=aes(x=Year, y = `Average Weekly Wages`, group = `Area Name`, color=`Area Name`)) +
#  labs(title = "Average Weekly Wages at Colleges and Universities in California", x  = "Year", y = "Average Weekly Wages") +
#  theme(legend.key.size = unit(0.2,"cm"), legend.spacing = unit(0.1,"cm")) +
#  facet_wrap(~Ownership) +
#  scale_color_manual(values = colors)
#ggplotly(p1)
colluniv_v2 <- adjust_for_inflation(colluniv$`Average Weekly Wages`, colluniv$Year, "US", to_date=2020)
colluniv_v3 <- cbind(colluniv, colluniv_v2)  
```

```{r geompoint, echo=FALSE, fig.width=8.5, fig.height=4}
p1 <- colluniv_v3[`Area Name` != "California"] %>% ggplot() +
  geom_point(mapping=aes(x=Year, y = `colluniv_v2`, group = `Area Name`, color=`Area Name`)) +
  labs(title = "Figure 1: Average Weekly Wages at Colleges and Universities in California", x  = "Year", y = "Average Weekly Wages") +
  #theme(legend.key.size = unit(0.2,"cm"), legend.spacing = unit(0.1,"cm")) +
  facet_wrap(~Ownership) +
  scale_color_manual(values = colors)
ggplotly(p1)
#p1
```

<sub><sup>**Figure 1** *Average Weekly Wages at Colleges and Universities in California* This graph shows the annual average weekly wage in the Colleges and Universities industry, separated by ownership (local government owned, privately owned, state government owned) throughout California. It is colored according to county.</sup></sub>
  
  To look more closely at how salary has grown, I calculated the percent changes (compared to the previous year reported) for privately owned and state government owned colleges and universities. 
  
  The bar graph plotting these percent changes for colleges and universities owned by the state government is below. Some years with more negative percent changes that stand out to me are 2008-2010, during which there was a recession.

```{r statepc, echo=FALSE}
#statecolluniv <- colluniv[Ownership == "State Government"] 
statecolluniv <- colluniv_v3[Ownership == "State Government"]
statecolluniv2 <- statecolluniv %>%
    group_by(`Area Name`) %>%
    arrange(Year) %>%
    mutate(pct.chg = 100 *(`colluniv_v2` - lag(`colluniv_v2`))/lag(`colluniv_v2`))
statecolluniv2 <- as.data.table(statecolluniv2)
p2 <- statecolluniv2[!is.na(pct.chg) & `Area Name` != "California"] %>%
  ggplot() +
  geom_bar(mapping = aes(x = Year, y = pct.chg, fill = `Area Name`), stat ="identity") +
  labs(title = "Figure 2: Percent Changes in Average Weekly Wages of State \n Government Owned Colleges and Universities in California", y = "Percent Change") +
  scale_fill_manual(values = colors) #+
  #theme(legend.key.size = unit(0.2,"cm"), legend.spacing = unit(0.1,"cm"))
ggplotly(p2)
#p2
```
<sub><sup>**Figure 2** *Percent Changes in Average Weekly Wages of State Government Owned Colleges and Universities in California* This graph shows the yearly percent change in the average weekly wages of State Government owned Colleges and Universities throughout California. It is colored according to county.</sup></sub>
  
  Below is the bar graph showing the percent changes for privately owned colleges and universities in California. Overall, I see more negative percent changes here compared to the previous graph, and this is likely due to the nature of these universities not having government regulation and having more autonomy with how they manage themselves. What stands out to me the most here is the positive percent change from 2018 to 2019 in San Mateo County. I believe Stanford University is the major private university in this county, and I found an article stating that their football coach David Shaw received "over 8.9 million in pay" in 2019, so this was likely a big contributor to the jump in average weekly wage for that year<sup>5</sup>.  
```{r privatepc, echo=FALSE}
privatecolluniv <- colluniv_v3[Ownership == "Private"] 
privatecolluniv2 <- privatecolluniv %>%
    group_by(`Area Name`) %>%
    arrange(Year) %>%
    mutate(pct.chg = 100 *(`colluniv_v2` - lag(`colluniv_v2`))/lag(`colluniv_v2`))
privatecolluniv2 <- as.data.table(privatecolluniv2)
p3 <- privatecolluniv2[!is.na(pct.chg) & `Area Name` != "California"] %>%
  ggplot() +
  geom_bar(mapping = aes(x = Year, y = pct.chg, fill = `Area Name`), stat ="identity") +
  labs(title = "Figure 3: Percent Changes in Average Weekly Wage of Privately \n Owned Colleges and Universities in California", y = "Percent Change") +
  scale_fill_manual(values = colors)# +
  #theme(legend.key.size = unit(0.2,"cm"), legend.spacing = unit(0.1,"cm"))
ggplotly(p3)
#p3
```
<sub><sup>**Figure 3** *Percent Changes in Average Weekly Wage of Privately Owned Colleges and Universities in California* This graph shows the yearly percent change in the average weekly wages of privately owned Colleges and Universities throughout California. It is colored according to county.</sup></sub>

# Conclusions
  From this analysis, I have found that counties with major cities generally have higher wages than those that do not. Many of the top wages come from industries that are tech-related and are especially prevalent in the areas surrounding Silicon Valley. There is less county-wide variation in average weekly wages in colleges and universities owned by the state government compared to those that are privately owned. Colleges and universities that are owned by the state government also appear to be more likely to have positive percent changes in annual wage.

# References
1. “NAICS &amp; Sic Identification Tools.” NAICS Association, https://www.naics.com/search/. 
2. “NAICS Code: 519190 All Other Information Services.” NAICS Association, https://www.naics.com/naics-code-description/?code=519190. 
3. “California Counties by Population.” California Outline, https://www.california-demographics.com/counties_by_population.
4. “Our Colleges.” LACCD Colleges, https://www.laccd.edu/about/pages/our-colleges.aspx. 
5. Berkowitz, Steve. “Stanford Football Coach David Shaw Credited with More than $8.9 Million in Pay for 2019.” USA Today, Gannett Satellite Information Network, 4 Aug. 2021, https://www.usatoday.com/story/sports/ncaaf/pac12/2021/08/04/stanford-football-coach-david-shaw-2019-pay/5488869001/. 

